Using numerical modelling to investigate temperature and wave exposure as drivers of macroalgal characteristics

### 2019-08-13 morphology_paper.Rmd
map added and word output updated
updated abiotic figures. theme_bw() will be used from now on. Extra details to figure added. All following graphs will be updated accordingly.

---
title: Using numerical modelling to investigate temperature and wave exposure as drivers of macroalgal characteristics
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
geometry: margin=1in
output:
  word_document:
    reference_docx: frontiers_template.docx
  html_document:
    df_print: paged
  pdf_document:
    fig_caption: yes
    fig_crop: no
    fig_height: 7
    fig_width: 7
  html_notebook:
    fig_caption: yes
    fig_height: 7
    fig_width: 7
papersize: A4
tables: yes
fontsize: 10pt
---

@@

@@

```{r schematic, echo=FALSE, fig.cap='Schematic of *L. pallida* and *E. maxima* morphology.'}
library(png)
library(grid)

@@

# Annual temp
temp_plot <- ggplot(data = temp_ann, aes(x = site, y = mean_temp)) +
  geom_point(show.legend = FALSE) + 
  scale_x_discrete(limits = positions_temp) + 
  ylim(9, 21) + 
  geom_errorbar(aes(x = site, ymin = mean_temp - sd_temp, ymax = mean_temp + sd_temp)) + 
  geom_point(data = temp_ann, aes(x = site, y = max_temp, colour = "blue"),show.legend = FALSE) + 
  geom_point(data = temp_ann, aes(x = site, y = min_temp, colour = "pink"),show.legend = FALSE) + 
  xlab("") + 
  ylab("") + 
  theme(plot.title = element_text(hjust = 0)) +
  theme(legend.position = "none") + 
  coord_flip()

@@

temp_boxplot <- ggplot(data = temp_ann_meta, aes(x = region, y = mean_temp, fill  = region)) + 
  geom_boxplot() + 
  stat_compare_means(label.y = 13) +
  xlab("") + 
  ylab("") +
  coord_flip()
```

```{r Temperature variables plot, echo=FALSE, warning=FALSE,fig.height=5,fig.width=6, fig.cap='Temperature variables at the collection sites around the Cape Peninsula. Temperature variables include minimum represented by blue dots, mean represented by balc dots, maximum by red dots and whiskers standard deviation.'}
ggarrange(temp_plot, temp_boxplot, common.legend = FALSE, ncol = 1, nrow = 2, labels = c("AUTO"), align = c("h"), legend = c("bottom"))
```

@@

```{r wave variable processing, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
### Plots for wave variables
#Annual wave plots
#combine with meta data for easier plotting
wave_ann_meta <- merge(meta_all, wave_ann, by = "site")
wave_ann_meta$site <- as.character(wave_ann_meta$site)#make character to ensure no issues arise for site and region names

positions_wave <- c("Betty's Bay", "Roman Rock","Kalk Bay", "A-Frame", "Miller's Point", "Baboon Rock",  "Batsata Rock","Bordjies Deep","Buffelsbaai","Buffelsbaai South","Olifantsbos","Scarborough", "Soetwater", "Kommetjie","Hout Bay","Oudekraal","Sea Point","St Helena Bay")

wave_plot_A <- ggplot(data = wave_ann, aes(x = site, y = hs_mean)) +
  geom_point() + 
  scale_x_discrete(limits = positions_wave) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5)) +
  geom_errorbar(aes(x = site, ymin = hs_mean - hs_sd, ymax = hs_mean + hs_sd)) + 
  xlab("") + 
  ylab("") + 
  theme(legend.position = "none") + 
  theme(plot.title = element_text(hjust = 0)) + 
  coord_flip()

wave_plot_B <- ggplot(data = wave_ann, aes(x = site, y = tp_mean)) +
  geom_point() + 
  scale_x_discrete(limits = positions_wave) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5)) +
  geom_errorbar(aes(x = site, ymin = tp_mean - tp_sd, ymax = tp_mean + tp_sd)) + 
  xlab("") + 
  ylab("") + 
  theme(legend.position = "none") + 
  theme(plot.title = element_text(hjust = 0)) + 
  coord_flip()

wave_boxplot_1 <- ggplot(data = wave_ann_meta, aes(x = region, y = hs_mean, fill  = region)) + 
  geom_boxplot() + 
  scale_fill_discrete(name="Region") +
  stat_compare_means(label.y =1,label.x = 1.5) + 
  xlab("") + 
  ylab("") + 
  coord_flip()

wave_boxplot_2 <- ggplot(data = wave_ann_meta, aes(x = region, y = tp_mean, fill  = region)) + 
  geom_boxplot() + 
  stat_compare_means(label.y = 9, label.x = 1.5) + 
  scale_fill_discrete(name="Region") +
  xlab("") + 
  ylab("") +
  ylim(8,12) + 
  coord_flip()

@@

```{r wave variable plot, warming = FALSE, echo=FALSE, message = FALSE, include= TRUE,fig.height=7,fig.width=7, fig.cap='Wave variables at each collection site around the Western Cape coast. Mean variables are represented by dots and standard deviation by whiskers.'}
ggarrange(wave_plot_A,wave_boxplot_1, wave_plot_B, wave_boxplot_2, common.legend = TRUE, labels = c("A1", "A2", "B1", "B2"))
```

@@

```{r wave rose figure, echo=FALSE, eval= FALSE, fig.cap='Wave rose figure'}

@@

We provide strong support for the idea that variations in the environmental variables, particularly wave exposure variables, are driving kelp morphological characteristics around the Cape Peninsula. Morphological adaptation due to water motion may manifest in a number of ways in high wave energy environments. For instance, reduction of blade thickness, blade elongation, increase of stipe length, increase in stipe circumference and force of attachment (Fowler-Walker et al. 2006; Wernberg & Thomsen 2005; Friedland & Denny 1995; Denny & Roberson 2002; Bekkby et al. 2014; Denny et al. 1997; Denny & Gaylord 2002) have been suggested in previous studies. Although this study did not measure force of attachment, other morphological responses to wave variables were evident. Higher temperatures tended to be assocaited with reduced measurements of stipe mass, stipe circumference/diameter and stipe length for the sub-canopy species *L. pallida*, which suggests that there is a reduction in overall surface area of the plant in warm temperate environments. The ecological consequence of this is unclear...The canopy species *E. maxima* exibits increased stipe length and decreased stipe circumference in warm temperate areas for both deep and shallow populations, which suggests a 'slender kelp' tactic and has been shown in previous work.**add examples here**. The response of kelp morphological characteristics to wave exposure was evident and both species show a tactics based on the magnitde of wave exposure. Strength increasing traits were exibited for lower exposure levels while a 'go-with-the-flow' tactic for moderate levels of wave exposure. Denny and Galord ??? showed that increased flexiblity or a 'go-with-the-flow' tactic has a threshold value, where increased stipe length actaully increase drag under high wave exposure scenarios. Therefore, under high wave exposure scenarios size reducing traits, such as reduced surface area of the overall plant should be favoured instead. This was evident in the results which showed that stipe length decreases under high wave exposure categories. This suggests that how morphological characteristcs manifest themselves is dependent on the magnitude of wave exposure. Furthermore, the results suggest that in the absence of wave exposure, temperatre becomes the dominant morphological driver. This is evident from the redundancy analysis for each species and deep populations of *E. maxima* which show clustering of study sites according to the region; wave variables drive morphological characteristics for kelps from the western region and temperature variables for kelps from the eastern region. Clustering of sites according to region overlapped almost completey for the RDA analysis of shallow *E. maxima* suggesting that for shallow kelp the effect of wave exposure is less variable.

@@

```{r appendix figure 1, echo=FALSE, message=FALSE, warning=FALSE}

@@

```{r appendix figure 2, echo=FALSE, message=FALSE, warning=FALSE}
# Order site names by location west to east

@@

ggarrange(eck_all_A,eck_all_B,eck_all_C,eck_all_D,eck_all_E,eck_all_F,eck_all_G,eck_all_H,eck_all_I, ncol = 3, nrow = 3, common.legend = TRUE)
```

>>>>

---
title: Using numerical modelling to investigate temperature and wave exposure as drivers of macroalgal characteristics
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
geometry: margin=1in
output:
  word_document:
    reference_docx: frontiers_template.docx
  pdf_document: 
  html_document:
    df_print: paged
  html_notebook:
    fig_caption: yes
    fig_height: 7
    fig_width: 7
papersize: A4
tables: yes
fontsize: 10pt
---

@@

```{r map study sites, echo=FALSE, message=FALSE, warning=FALSE}
library(png)
library(grid)

map <- readPNG("figures/sa_map.png")
grid.raster(map)
```

@@

```{r schematic, echo=FALSE, fig.cap='Schematic of *L. pallida* and *E. maxima* morphology.'}

@@

# Annual temp
temp_plot <- ggplot(data = temp_ann, aes(x = site, y = mean_temp)) +
  geom_point(show.legend = FALSE) + 
  scale_x_discrete(limits = positions_temp) + 
  ylim(9, 21) + 
  geom_errorbar(aes(x = site, ymin = mean_temp - sd_temp, ymax = mean_temp + sd_temp)) + 
  geom_point(data = temp_ann, aes(x = site, y = max_temp, colour = "blue"),show.legend = FALSE) + 
  geom_point(data = temp_ann, aes(x = site, y = min_temp, colour = "pink"),show.legend = FALSE) + 
  xlab("") + 
  ylab("") + 
  theme(plot.title = element_text(hjust = 0)) +
  theme(legend.position = "none") + 
  coord_flip() + 
  theme_bw()

@@

temp_boxplot <- ggplot(data = temp_ann_meta, aes(x = region, y = mean_temp, fill  = region)) + 
  geom_boxplot() + 
  stat_compare_means(label.y = 14.5 ,label.x = 1.5, label = "p.format") +
  xlab("") + 
  ylab("") +
  coord_flip() + 
  theme_bw()

```{r Temperature variables plot, echo=FALSE, warning=FALSE,fig.height=5,fig.width=8, fig.cap='Temperature variables at the collection sites around the Cape Peninsula are represented in plot A. Temperature variables include minimum represented by blue dots, mean represented by balc dots, maximum by red dots and whiskers standard deviation. Summary of temperature data for sites grouped by region are represented in plot B. See appendix for significance level guide.'}
ggarrange(temp_plot, temp_boxplot, common.legend = FALSE, ncol = 2, nrow = 1, labels = c("AUTO"), align = c("h"), legend = FALSE)
```

@@

```{r wave variable processing, message=FALSE, warning=FALSE, include=FALSE}
### Plots for wave variables
#Annual wave plots
#combine with meta data for easier plotting
wave_ann_meta <- merge(meta_all, wave_ann, by = "site")
wave_ann_meta$site <- as.character(wave_ann_meta$site)#make character to ensure no issues arise for site and region names

positions_wave <- c("Betty's Bay", "Roman Rock","Kalk Bay", "A-Frame", "Miller's Point", "Baboon Rock",  "Batsata Rock","Bordjies Deep","Buffelsbaai","Buffelsbaai South","Olifantsbos","Scarborough", "Soetwater", "Kommetjie","Hout Bay","Oudekraal","Sea Point","St Helena Bay")
cols <- c("Betty's Bay" = "red", "Roman Rock"= "red","Kalk Bay"= "red", "A-Frame"= "red", "Miller's Point"= "red", "Baboon Rock"= "red",  "Batsata Rock"= "red","Bordjies Deep"= "red","Buffelsbaai"= "red","Buffelsbaai South"= "red","Olifantsbos"= "blue","Scarborough"= "blue", "Soetwater"= "blue", "Kommetjie"= "blue","Hout Bay"= "blue","Oudekraal"= "blue","Sea Point"= "blue","St Helena Bay"= "blue")

wave_plot_A <- ggplot(data = abio_clim, aes(x = site, y = hs_mean)) +
  geom_point(aes(colour = factor(site)), show.legend = FALSE) + 
  scale_x_discrete(limits = positions_wave) + 
  scale_colour_manual(values = cols) +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5)) +
  geom_errorbar(aes(x = site, ymin = hs_mean - hs_sd, ymax = hs_mean + hs_sd)) + 
  xlab("") + 
  ylab("") + 
  theme(legend.position = "none") + 
  theme(plot.title = element_text(hjust = 0)) + 
  coord_flip() + 
  theme_bw()

wave_plot_B <- ggplot(data = abio_clim, aes(x = site, y = tp_mean)) +
  geom_point(aes(colour = factor(site)), show.legend = FALSE) + 
  scale_x_discrete(limits = positions_wave) + 
  scale_colour_manual(values = cols) +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5)) +
  geom_errorbar(aes(x = site, ymin = tp_mean - tp_sd, ymax = tp_mean + tp_sd)) + 
  xlab("") + 
  ylab("") + 
  theme(legend.position = "none") + 
  theme(plot.title = element_text(hjust = 0)) + 
  coord_flip() +
  theme_bw()

wave_boxplot_1 <- ggplot(data = wave_ann_meta, aes(x = region, y = hs_mean, fill  = region)) + 
  geom_boxplot(show.legend = FALSE) + 
  stat_compare_means(label.y =1.4,label.x = 1.5, label = "p.format") + 
  xlab("") + 
  ylab("") + 
  coord_flip() + 
  theme_bw()

wave_boxplot_2 <- ggplot(data = wave_ann_meta, aes(x = region, y = tp_mean, fill  = region)) + 
  geom_boxplot(show.legend = FALSE) + 
  stat_compare_means(label.y = 10, label.x = 1.5, label = "p.format") + 
  xlab("") + 
  ylab("") +
  ylim(8,12) + 
  coord_flip() + 
  theme_bw()

@@

```{r wave variable plot, warning = FALSE, echo=FALSE, message = FALSE, include= TRUE,fig.height=7,fig.width=8, fig.cap='In plot A1 and A2 wave variables at each collection site around the Western Cape coast are presented. Mean variables are represented by dots and standard deviation by whiskers. Blue dots represent sites grouped to the western category and red dots represent sites grouped to the eastern category.'}
ggarrange(wave_plot_A,wave_boxplot_1, wave_plot_B, wave_boxplot_2, common.legend = TRUE, labels = c("A1", "A2", "B1", "B2"))
```

@@

```{r wave rose figure, echo=FALSE, warning = FALSE, eval= FALSE, fig.cap='Wave rose figure'}

@@

We provide strong support that variations in the environmental variables, particularly wave exposure variables, are driving kelp morphological characteristics around the Cape Peninsula. Morphological adaptation due to water motion may manifest in a number of ways in high wave energy environments. For instance, reduction of blade thickness, blade elongation, increase of stipe length, increase in stipe circumference and force of attachment (Fowler-Walker et al. 2006; Wernberg & Thomsen 2005; Friedland & Denny 1995; Denny & Roberson 2002; Bekkby et al. 2014; Denny et al. 1997; Denny & Gaylord 2002) have been suggested in previous studies. Although this study did not measure force of attachment, other morphological responses to wave variables were evident. Higher temperatures tended to be assocaited with reduced measurements of stipe mass, stipe circumference/diameter and stipe length for the sub-canopy species *L. pallida*, which suggests that there is a reduction in overall surface area of the plant in warm temperate environments. The ecological consequence of this is unclear...The canopy species *E. maxima* exibits increased stipe length and decreased stipe circumference in warm temperate areas for both deep and shallow populations, which suggests a 'slender kelp' tactic and has been shown in previous work.**add examples here**. The response of kelp morphological characteristics to wave exposure was evident and both species show a tactics based on the magnitde of wave exposure. Strength increasing traits were exibited for lower exposure levels while a 'go-with-the-flow' tactic for moderate levels of wave exposure. Denny and Galord ??? showed that increased flexiblity or a 'go-with-the-flow' tactic has a threshold value, where increased stipe length actaully increase drag under high wave exposure scenarios. Therefore, under high wave exposure scenarios size reducing traits, such as reduced surface area of the overall plant should be favoured instead. This was evident in the results which showed that stipe length decreases under high wave exposure categories. This suggests that how morphological characteristcs manifest themselves is dependent on the magnitude of wave exposure. Furthermore, the results suggest that in the absence of wave exposure, temperatre becomes the dominant morphological driver. This is evident from the redundancy analysis for each species and deep populations of *E. maxima* which show clustering of study sites according to the region; wave variables drive morphological characteristics for kelps from the western region and temperature variables for kelps from the eastern region. Clustering of sites according to region overlapped almost completey for the RDA analysis of shallow *E. maxima* suggesting that for shallow kelp the effect of wave exposure is less variable.

@@

```{r appendix figure 1, echo=FALSE, message=FALSE, warning=FALSE, fig.height=7,fig.width=7}

@@

```{r appendix figure 2, echo=FALSE, message=FALSE, warning=FALSE, fig.height=7,fig.width=7}
# Order site names by location west to east

@@

ggarrange(eck_all_A,eck_all_B,eck_all_C,eck_all_D,eck_all_E,eck_all_F,eck_all_G,eck_all_H,eck_all_I, ncol = 3, nrow = 3, common.legend = TRUE, legend = FALSE)
```
